%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter5.tex
%% NOVA thesis document file
%%
%% Chapter with lots of dummy text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter5.tex}%

\chapter{Work Plan}
\label{cha:work_plan}

\section{Gantt Chart}
\label{section:gantt_chart}

\begin{itemize}
    \setlength\itemsep{0bp}
    \item{\emph{Task 1: Research}}\hspace{1em}
    For the month of March, we will be researching study cases and analysing them in 
    \ortac + \cameleer. This will provide data for the next tasks throughout the work.
    \item{\emph{Task 2: Define E-Gospel}}\hspace{1em}
    The next step will be to define E-Gospel - an executable portion of both \ocaml and 
    specification/contract tools that can be used in \gospellang. The defining of E-Gospel 
    will use the data from the research and analysis. This will be done throughout April 
    until the middle of May.
    \item{\emph{Task 3: Define Monitors}}\hspace{1em}
    Using the previous steps as a base, we will define the specification of monitors that 
    are best suited according the conclusions of the previous tasks. This will be done 
    throughout May until the end of June.
    \item{\emph{Task 4: Implement Monitors}}\hspace{1em}
    From mid June throughout July and August, we will implement the monitors previously defined 
    and integrate the first set of data to test them. It is expected that the first 
    implementation will be ready by the end of August.
    \item{\emph{Task 5: Writing}}\hspace{1em}
    The last month of the work will be dedicated to writing the final document. This will
    occupy the month of September.
\end{itemize}

\begin{figure}[h]
    \tikzset{every picture/.style={xscale=0.75,yscale=1,transform shape}}
    \begin{ganttchart}[ y unit chart = 0.6cm,
                        vgrid,
                        bar top shift=-0.1,
                        bar height=0.6,
                        title height=0.7]{1}{28}
      \gantttitle{March}{4}
      \gantttitle{April}{4}
      \gantttitle{May}{4}
      \gantttitle{June}{4}
      \gantttitle{July}{4}
      \gantttitle{August}{4}
      \gantttitle{September}{4}\\
      \gantttitlelist{1,...,4}{1}
      \gantttitlelist{5,...,8}{1}
      \gantttitlelist{9,...,12}{1}
      \gantttitlelist{13,...,16}{1}
      \gantttitlelist{17,...,20}{1}
      \gantttitlelist{21,...,24}{1}
      \gantttitlelist{25,...,28}{1} \\
    
    \ganttbar{Research \& Analysis}{1}{4} \\
    
    \ganttbar{Define E-Gospel}{5}{10} \\
    
    \ganttbar{Define Monitors}{9}{16} \\
    
    \ganttbar{Implement Monitors}{15}{24} \\
    
    \ganttbar{Writing}{25}{28}
        
    \end{ganttchart}
    \caption{Planned Schedule}
    \label{ganttchart}
    
\end{figure}