%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% glossary.tex
%% NOVA thesis document file
%%
%% Glossary definition
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE glossary.tex}%

\newglossaryentry{rac}{
  name={RAC},
  sort={rac},
  description={Runtime Assertion Checking - a specific technique that resorts to assertions to check the correctness of a program during its execution.}
}
 
\newglossaryentry{dv}{
  name={DV},
  sort={dv},
  description={Dynamic Verification - process of checking the correctness of a program during its execution.}
}

\newglossaryentry{gospel}{
  name={GOSPEL},
  sort={gospel},
  description={Generic OCaml SPEcification Language - a behavioral specification language for OCaml.}
}

\newglossaryentry{cameleer}{
  name={Cameleer},
  sort={cameleer},
  description={A deductive verification tool for OCaml programs.}
}

\newglossaryentry{why3}{
  name={Why3},
  sort={why3},
  description={A platform for deductive program verification.}
}

\newglossaryentry{ortac}{
  name={ORTAC},
  sort={ortac},
  description={OCaml Runtime Assertion Checker - a tool to integrate runtime assertion checking into OCaml programs.}
}

\newglossaryentry{z3}{
  name={Z3},
  sort={z3},
  description={An efficient SMT solver developed at Microsoft Research.}
}

\newglossaryentry{altergo}{
  name={Alt-Ergo},
  sort={alt-ergo},
  description={An open-source SMT solver developed at LRI, INRIA.}
}

\newglossaryentry{coq}{
  name={Coq},
  sort={coq},
  description={A formal proof management system.}
}